---
title: "Soil Moisture"
author: "Andrew"
date: "10/7/2019"
output:
  html_document:
    toc: true
    toc_depth: 2

external_link: ""
image:
  caption: ''
  focal_point: Smart
links:
- icon: twitter
  icon_pack: fab
  name: Follow
  url: https://twitter.com/A_MURRAY89
slides: 
summary: Integral to my work in identifying areas vulnerable to groundwater contamination following extreme precipitation events.
tags:
- Water
- Soil Moisture
- Hurricanes
url_code: ""
url_pdf: ""
url_slides: "/slides/soil_moisture/SM_slides.html"
url_video: ""
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(soiltexture)
```

## Soil Moisture and Groundwater Vulnerability

Soil moisture is an important indicator of the state of the hydrologic cycle. Typically, soil moisture is thought of as an important indicator for how well vegetation is doing such as in major agriculturl areas. For example, it doesn't matter how much it rained yesterday if the soil could not retain any of it. Whereas precipitation tells informs us of how much water is falling to the ground, soil moisture informs us how much water is available for uptake by vegetation. Related to this is the landscapes capacity to absorb water. 

```{r soilTexture, echo=FALSE, fig.cap= "USDA Soil Texture Triangle"}
TT.plot( class.sys = "USDA.TT" )

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
